Community-Conditioned Language Models

### 2020-11-22 paper/paper.tex
fix information gain computation/definiton

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{latexsym}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{array}
\usepackage{cleveref}

\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.16}
\pgfplotstableset{
  column type=r,
  fixed zerofill,
  clear infinite,
  %dec sep align,
  %empty cells with={\ensuremath{-}},
}

@@

\newcommand\Ppl{\mathsf{Ppl}}
\newcommand\Ind{I}
\newcommand\BibTeX{B\textsc{ib}\TeX}

@@

We trained the models until the validation loss stopped decreasing 
for two epochs in a row,
and used the weights from the epoch with the smallest validation loss for testing.
In general, transformer models trained for about half as long as the LSTM models (\cref{tab:best-epoch}).

\begin{table}
  \centering
  \input{floats/best_epoch.tex}
  \caption{Epoch with the lowest validation loss.}
  \label{tab:best-epoch}
\end{table}

%\pgfplotstabletypeset[
  %columns={best_epoch,lm_ppl,gain},
  %columns/model/.style={string type,column type=l},
  %columns/best_epoch/.style={column name={Epochs}, int detect},
  %columns/lm_ppl/.style={column name={Ppl}},
  %columns/gain/.style={column name={IG},precision=3,empty cells with={\ensuremath{-}}},
  %every nth row={5}{before row=\midrule},
  %]{floats/model.csv}

@@

We measure the perplexity of each example in the test set. 
For this, we use the negative exponential of the mean cross-entropy loss 
for the sequence.

As one would expect, 
the conditioned models have lower perplexity
than their respective unconditioned baseline models,
but this effect is different in magnitude
for models with different community embedding depths.
For the LSTM encoder, 
the best model ($l_c=1$) concatenates the community embedding after the first encoder layer,
with two more LSTM layers after that.
The best transformer model ($l_c=3$), on the other hand,
incorporates the community information last,
after all the sequence encoder layers
(\cref{tab:model-ppls}).

%TODO: For space reasons, it might be better to put the community-level table in the appendix and have the table in body just show the means (and perhaps medians/stddevs?) for each architecture.}

We also note that the conditioned models perform differently across different communities---%
for some communities, perplexity of the conditioned models is a lot lower than the unconditioned models,
whereas for others it is the same or even slightly higher (\cref{tab:model-ppls}).
\bn{talk about this in terms of mutual information instead!}
As an informal observation, we note 
that across all the models we tested, 
communities where conditioning has the least effect
tend to be more organized around more general interest topics,
such as the \emph{relationships} subreddit, where the subject matter is relevant to a broad
range of people.
Conditioning the model on community has the most benefit
for narrower special-interest subreddits such as those organized
around a certain videogame, sports team, or subculture.
This makes sense intuitively, since communities with more niche topics
would tend to have more specialized vocabulary and other linguistic patterns.

\begin{table*}
  \footnotesize
  \centering
  \input{floats/model_ppls.tex}
  \caption{Mean model perplexity for messages in the test set, broken down by community.}
  \label{tab:model-ppls}
\end{table*}

@@

We consider the average information gain of each CCLM over its baseline
un-conditioned counterpart with the same sequence encoder architecture.
For a given message, the information gain is defined as the ratio of the
perplexity of the baseline model to that of the CCLM:
\[
  IG_{CCLM}(m) = \frac{Ppl_{LM}(m)}{Ppl_{CCLM}(m)}
\]
\jp{What we had before: average entropy gain is $H(m)-H_0(m)$, and
  then taking the average over all messages:
  $avg(H(m)-H_0(m)) = avg(H(m))-avg(H_0(m)) = H - H₀$. Taking the exponential
  gives a ratio of perplexity.  Here it seems that you have commuted
  exponential and average. It is rather weird to do average in the
  exponential space without normalising to a probability.

  Compare: $e^{avg(H(m))} / e^{avg(H_0(m))}$ with $avg \frac{e^{H(m)}}{e^{H₀(m)}}$.

}
\begin{table*}
  \centering
  % \input{floats/info_gain.tex}
  \caption{Mean information gain of the CCLM over the baseline model, for messages in the test set.}
  \label{tab:model-info-gain}
\end{table*}

While the absolute performance of the LSTM models is better,
we find the the transformer CCLMs have higher information gain 
\cref{tab:model-info-gain}.

@@

As a qualitative assesment of the community embeddings, 
we manually identified topic categories into which at least three
of the selected communities fall.
We plotted the two-component PCA projection of the embeddings,
annotated with these categories (\cref{fig:comm-pca}).
We see that there is some evidence of clustering according to
the topic categories we identified,
suggesting that the embedding space learned by the model is meaningful.

@@

\begin{figure*}
\begin{tikzpicture}
  \confusiongroupplot{lstm-3-1}{transformer-3-3}
\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{cleveref}
\usepackage{array}
\usepackage{makecell}
\usepackage{multirow}

\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.16}

@@

\newcommand\Ppl{\mathsf{Ppl}}
\newcommand\IG{\mathsf{IG}}
\newcommand\Ind{I}
\DeclareMathOperator*{\avg}{avg}
\newcommand\BibTeX{B\textsc{ib}\TeX}

@@

We trained the models until the validation loss stopped decreasing 
for two epochs in a row,
and used the weights from the epoch with the smallest validation loss for testing.
In general, transformer models trained for about half as long as the LSTM models (test epoch, \cref{tab:model-results}).

@@

We measure the perplexity of each example in the test set. 
For this, we use the negative exponential of the mean cross-entropy loss 
of the language model for a given message and average over a set of messages, $M$.
\[\Ppl_M = \avg_{m\in M}(-e^{H(m)})\] \bn{Following the logic from IG, should we change Ppl to the exponential of the average cross entropy, rather than using the average of the message-level perplexity?} 
As one would expect, 
the conditioned models have lower perplexity
than their respective unconditioned baseline models (\cref{tab:model-results}),
but this effect is different in magnitude
for models with different community embedding depths.
For the LSTM encoder, 
the best model ($l_c=1$) concatenates the community embedding after the first encoder layer,
with two subsequent LSTM layers. 
The best transformer model ($l_c=3$), on the other hand,
incorporates the community information last,
after all the sequence encoder layers.

\begin{table}
  \footnotesize
  \centering
  \input{floats/model_results.tex}
  \caption{
    Performance of baseline (first row for each encoder architecture) and CCLM models. 
    Perplexity and information gain are shown for the entire test set 
    (\num{5000} messages for each community).}
  \label{tab:model-results}
\end{table}

We also consider model performance stratified by community;
that is, $\Ppl_{M_j}$, where $M_j$ is the test-set messages from community $c_j$.
We observe a great deal of variation in baseline perplexity across communities,
with $\Ppl_{M_j}$ ranging from \num{53.07} to \num{115.29} for the LSTM model
(\cref{tab:community-results}).

@@

Next, we consider the \emph{information gain} of each CCLM over its baseline
un-conditioned counterpart with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{LM}(m) - H_{CCLM}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_M(H_{LM}(m))}}{e^{\avg_M(H_{CCLM}(m))}}\]
%\jp{What we had before: average entropy gain is $H(m)-H_0(m)$, and
  %then taking the average over all messages:
  %$avg(H(m)-H_0(m)) = avg(H(m))-avg(H_0(m)) = H - H₀$. Taking the exponential
  %gives a ratio of perplexity.  Here it seems that you have commuted
  %exponential and average. It is rather weird to do average in the
  %exponential space without normalising to a probability.

  %Compare: $e^{avg(H(m))} / e^{avg(H_0(m))}$ with $avg \frac{e^{H(m)}}{e^{H₀(m)}}$.
%}

While the absolute performance of the LSTM models is better,
we find that the best transformer CCLM has higher information gain 
than the best LSTM (\cref{tab:model-results}).

The conditioned models also perform differently across different communities---%
for some communities, $\IG_{M_j}$ around \num{1}, or even slightly lower,
meaning that the CCLM has no advantage over the unconditioned baseline for that community.
For others communities it is higher, meaning that the CCLM performs better than 
the baseline for those communities (\cref{tab:community-results}).

\begin{table*}
  \normalsize
  \centering
  \input{floats/community_results.tex}
  \caption{
    Model performance by community. We show the absolute perplexity 
    of the baseline models and the information gain of the CCLMs.
    Note that the rows are sorted by the perplexity of the baseline
    LSTM model.
}
  \label{tab:community-results}
\end{table*}

As an informal observation, we note 
that across all the models we tested, 
communities where conditioning has the least effect
tend to be more organized around more general interest topics,
such as the \emph{relationships} and \emph{advice} subreddits, where the subject matter is relevant to a broad
range of people.
Conditioning the model on community has the most benefit
for narrower special-interest subreddits, such as those organized
around a certain videogame, sports team, or subculture.
This makes sense intuitively, since communities with more niche topics
would tend to have more specialized vocabulary and other linguistic patterns.

@@

As a qualitative assessment of the community embeddings, 
we manually identified topic categories into which at least three
of the selected communities fall.
We plotted the two-component PCA projection of the embeddings,
annotated with these categories (\cref{fig:comm-pca}).
We see that there is some evidence of clustering according to
the topic categories we identified,
suggesting that the embedding space learned by the model is meaningful.

@@

\begin{figure*}
%\begin{tikzpicture}
  %\confusiongroupplot{lstm-3-1}{transformer-3-3}
%\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}
